---
title: "Project Write-up"
format: html
editor: visual
---

# A Look Into Salaries in Data Science

## Introduction:

  As different career paths fall in and out of demand, it is hard to know what is a worth while pursuit in today's job market. We are all interested in careers in data science and we want be knowledgeable about the field as we prepare for graduation. We were especially curious about two things:
-  How do company-related factors influence the salary of workers in the field of data science??
-  How do personal factors (like location, experience level, and employment type) influence the salary of workers in the field of data science?
  To answer these questions we first selected a data set titled "Jobs and Salaries in Data Science," which originated from [AI Jobs](https://ai-jobs.net/salaries/2023/). Each observation represents one person in one year. Our dataset contains 9355 observations with 12 variables. The variables include information on salary, and explanatory variables such as company size, experience level, work setting, employment type, and year. There is also an extensive list of data science job titles.

```{r, echo=FALSE}
libs <- c('tidyverse','knitr','viridis', 'mosaic', 'Lahman', "readr", "dplyr", "tidyr", "ggplot2")
for(l in libs){
  if(!require(l,character.only = TRUE, quietly = TRUE)){
    message( sprintf('Did not have the required package << %s >> installed. Downloading now ... ',l))
    install.packages(l) 
  }
  library(l, character.only = TRUE, quietly = TRUE)
}
```

```{r, echo=FALSE}
data_science <- read_csv(file = "data/jobs_in_data.csv", show_col_types = FALSE)
```

```{r echo = F}
# Data wrangling
data_science$job_title <- as.factor(data_science$job_title)
data_science$experience_level <- as.factor(data_science$experience_level)
data_science$employment_type <- as.factor(data_science$employment_type)
data_science$work_setting <- as.factor(data_science$work_setting)
data_science$company_size <- as.factor(data_science$company_size)
```

### The Effect of Company-Level Demands and Characteristics on Salary

#### Introduction:

  In this section, we aim to answer the question 'how do company-related factors (like job title, job category, work setting, and company size) influence the salary of workers in the field of data science?' In order to get a complete understanding of data science salaries, it is important to disaggregate this data such that we are able to extract more specific insights. 
  We decided looking at average salaries by job title, job category, work setting, and company size provide a useful glimpse into the relative *value* (in a strictly monetary sense) of certain jobs and certain company attributes. First, viewing salary by job title conveys important information about which specific jobs are in high demand (or low supply). Job category is useful in a similar way, as it provides a more zoomed-out view of the data science landscape. Since the COVID pandemic increased the need and acceptability of remote work, we thought it would be appropriate to investigate the differences in salary by work environment. Finally, company size provides a useful understand of how differently sized companies pay their data science employees.

#### Approach (1-2 paragraphs):

  In order to provide clear insights about salary by company-level characteristics with minimal complexities and the primary intention of communicating the *makeup* of the data science field, we concluded that bar plots would make the most effective visualizations. The utility of bar plots is particularly handy for the scope of this project since many of the variables we are investigating in relation to salary are discrete, categorical variables.
  In particular, our first visualization, combines two bar plots: one relating to average salary by job title and the other relating to average salary by job category. As discussed briefly above, these plots convey similar information but with different levels of specificity, making the combined plot a reasonable glimpse at the sub-fields of data science. Our second visualization relates more to company characteristics than it does to fields/positions that are in demand by companies. Instead, this combined plot displays the relationship between work setting and company size, providing more environment-specific insights.


#### Analysis (2-3 code blocks, 2 figures, text/code comments as needed):


```{r, echo = FALSE}
library(cowplot)
# store the number of each kind of job title
job_title_counts <- data_science %>%
  count(job_title)

# filters out the jobs that appear fewer than 100 times
other_jobs <- job_title_counts %>%
  filter(n < 100) %>%
  pull(job_title)

# Assigns 'Other' to the jobs that appear fewer than 100 times
data_science_grouped <- data_science %>%
  mutate(job_title_grouped = if_else(job_title %in% other_jobs, "Other", job_title))

# Calculate the average salary for each job title
avg_salary_by_job <- data_science_grouped %>%
  group_by(job_title_grouped) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))

# Plotting the average salary for each job title by year
plot6 <- ggplot(data = avg_salary_by_job, aes(x = job_title_grouped, y = avg_salary)) +
  geom_bar(stat = "identity", fill = "lightpink") +
  theme_minimal() +
  labs(title = "Average Salary by Job Title",
       x = "Job Title",
       y = "Average Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

job_category_counts <- data_science %>%
  count(job_category)
other_jobs_C <- job_category_counts %>%
  filter(n < 100) %>%
  pull(job_category)

data_science_grouped_C <- data_science %>%
  mutate(job_category_grouped_C = if_else(job_category %in% other_jobs_C, "Other", job_category))
avg_salary_by_job_C <- data_science_grouped_C %>%
  group_by(job_category_grouped_C) %>%
  summarize(avg_salary = mean(salary, na.rm = TRUE))

plot7<- ggplot(data = avg_salary_by_job_C, aes(x = job_category_grouped_C, y = avg_salary)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Average Salary by Job Category",
       x = "Job Category",
       y = "Average Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

combined_plot3 <- plot_grid(plot6,plot7,ncol = 2)
combined_plot3
```

```{r}
library(cowplot)

# summary statistics of average salary by work setting
work_salary <- data_science %>%
  group_by(work_setting) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))

# summary statistics of average salary by company size
company_salary <- data_science %>%
  group_by(company_size) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))


# Plot 3: Average Salary by Work Setting
plot1 <- ggplot(data = work_salary, aes(x = work_setting, y = avg_salary)) +
  geom_bar(stat = "identity", fill = "lightpink") +
  theme_minimal() +
  labs(title = "Average Salary by Work Setting",
       x = "Work Setting",
       y = "Average Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot 4: Average Salary by Company Size
plot2 <- ggplot(data = company_salary, aes(x = factor(company_size, levels = c("S", "M", "L")), y = avg_salary)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  theme_minimal() +
  labs(title = "Average Salary by Company Size",
       x = "Company Size",
       y = "Average Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = c("S" = "Small", "M" = "Medium", "L" = "Large"))

# Arrange the plots in a grid
combined_plot <- plot_grid(plot1, plot2, ncol = 2)

# Print the combined plot
combined_plot
```


#### Discussion (1-3 paragraphs):

In the Discussion section, interpret the results of your analysis. Identify any trends revealed (or not revealed) by the plots. Speculate about why the data looks the way it does.

### The Effect of Personal Characteristics on Salary

#### Introduction

  In this section we aim to answer the question 'how do personal factors (like company location, experience level, and employment type) influence the salary of workers in the field of data science?' We believe that these variables provide a breadth of information about the landscape of data science salaries based on person factors.
  Let's look at each of these variables individually. First, while one cannot immediately change their level of experience in a given field, it is useful to understand expected pay at different portions of one's career. On the other hand, often people do have some say in their employment type and may decide to pursue certain employment types based on their average salaries. Next, the relationship between company location and salary is particularly interesting as data science is a relatively new field and requires a lot of technological infrastructure to perform. In addition, in many developing countries, there is likely not as high of a demand for data scientist. As such, we are interested in the differential pay between countries for similar jobs.
  

#### Approach

  Similar to our approach for investigating company-level characteristics, since our variables are categorical, it is reasonable to provide bar plots to display relationships between these categorical variables are salary. In addition, since we are also interested in location, a map provides an intuitive understand of differences in pay between different countries.
  First, we decided to combine two bar plots: one displaying average salary by experience level and the other displaying average salary by employment type. Next, we analyze the effect of company location on average salary. Note that while we are analyzing company location and not country of origin, we believe that workers' access to jobs *usually* depends on where they live, so that looking into company location still produces personal insights. 

#### Analysis

```{r}
library(cowplot)

# summary statistics of average salary by experience level
exp_salary <- data_science %>%
  group_by(experience_level) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))

# summary statistics of average salary by employment type
employment_salary <- data_science %>%
  group_by(employment_type) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))

# Plot 1: Average Salary by Experience Level
plot3 <- ggplot(data = exp_salary, aes(x = experience_level, y = avg_salary)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  theme_minimal() +
  labs(title = "Average Salary by Experience Level",
       x = "Experience Level",
       y = "Avg Salary") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Plot 2: Average Salary by Employment Type
plot4 <- ggplot(data = employment_salary, aes(x = employment_type, y = avg_salary)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  theme_minimal() +
  labs(title = "Average Salary by Employment Type",
       x = "Employment Type",
       y = "Avg Salary") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Arrange the plots in a grid
combined_plot2 <- plot_grid(plot3, plot4, ncol = 2)

# Print the combined plot
combined_plot2
```

```{r}
library(tmap)

avg_salary_by_country <- data_science %>%
  group_by(company_location) %>%
  summarize(avg_salary_in_usd = mean(salary_in_usd))

avg_salary_by_country <- avg_salary_by_country %>%
  mutate(company_location = case_when(
    company_location == "Czech Republic" ~ "Czech Rep.",
    company_location == "South Korea" ~ "Korea",
    company_location == "Central African Republic" ~ "Central African Rep.",
    company_location == "Bosnia and Herzegovina" ~ "Bosnia and Herz.",
    TRUE ~ company_location
  ))


data("World")
world_map <- merge(World, avg_salary_by_country, by.x = "name", by.y = "company_location", all.x = TRUE)

tm_shape(world_map) +
  tm_polygons(col = "avg_salary_in_usd", style = "jenks", palette = "Blues",
              main.title = "Average Salaries of Data Scientists by Country, 2020-2023",
              title = "Average Salaries by Country (USD), 2020-2023", legend.title = "Salary Range")
```

#### Discussion

  As displayed in the map, it appears that data scientists in highly developed nations tend to be compensate more than those in developing countries. Namely, companies in the United States, Canada, Saudi Arabia, Australia, and Japan pay their data scientists the most, on average, based on this data. On the other hand, Indonesia, Malaysia, Egypt, Turkey, Thailand, Pakistan, and Ecuador pay the least on average. It is important to note, however, that many countries, in particular many countries in Africa, South America, and the Eurasian Steppes have no data in this dataset on data scientist salaries. This is likely because most of these countries have very few data scientists, but could also be caused simply by a lack of access to salary data in these countries. In addition, this visualization doesn't take into account cost of living in these different countries.
  