---
title: "Grab the Cash: Money in Data Science"
subtitle: "Elliott Chang, Cameron Adams, Zoe Drajem"
format: 
  revealjs
highlight-style: a11y
execute:
  code-link: true
  echo: true
editor_options: 
  chunk_output_type: console
---

## Introduction {.smaller}

```{r, echo=FALSE}
libs <- c('tidyverse','knitr','viridis', 'mosaic', 'Lahman', "readr", "dplyr", "tidyr", "ggplot2", "tmap", "cowplot")
for(l in libs){
  if(!require(l,character.only = TRUE, quietly = TRUE)){
    message( sprintf('Did not have the required package << %s >> installed. Downloading now ... ',l))
    install.packages(l) 
  }
  library(l, character.only = TRUE, quietly = TRUE)
}
```

```{r, echo=FALSE}
data_science <- read_csv(file = "data/jobs_in_data.csv", show_col_types = FALSE)
```

As different career paths fall in and out of demand, it is hard to know what is a worthwhile pursuit in today's job market. We are all interested in careers in data science and we want be knowledgeable about the field as we prepare for graduation.

Our data:

-   Variables: salary (in local currency), salary (in USD), job title, job category, country of origin, company location (country), work year, work experience, age, work setting, company size, and employment type
-   Observations represent individual employees in a given year (2020-2023)
-   9355 rows, 12 columns
-   Source: https://ai-jobs.net/salaries/2023/


Q1: How do company-related factors (like job title, job category,work setting, and company size) influence the salary of workers in the field of data science?

Q2: How do personal factors (like location, experience level, and employment type) influence the salary of workers in the field of data science?

## The Effect of Company-Level Demands and Characteristics on Salary

```{r echo = F}
# Data wrangling
data_science$job_title <- as.factor(data_science$job_title)
data_science$experience_level <- as.factor(data_science$experience_level)
data_science$employment_type <- as.factor(data_science$employment_type)
data_science$work_setting <- as.factor(data_science$work_setting)
data_science$company_size <- as.factor(data_science$company_size)
```

```{r, echo = FALSE}
library(cowplot)
# store the number of each kind of job title
job_title_counts <- data_science %>%
  count(job_title)

# filters out the jobs that appear fewer than 100 times
other_jobs <- job_title_counts %>%
  filter(n < 100) %>%
  pull(job_title)

# Assigns 'Other' to the jobs that appear fewer than 100 times
data_science_grouped <- data_science %>%
  mutate(job_title_grouped = if_else(job_title %in% other_jobs, "Other", job_title))

# Plotting the average salary for each job title by year
plot6 <- ggplot(data = data_science_grouped, aes(x = job_title_grouped, y = salary_in_usd, fill = job_title_grouped)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Data Science Salaries by Job Title",
       x = "Job Title",
       y = "Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")+
  scale_y_continuous(labels = scales::label_comma())

job_category_counts <- data_science %>%
  count(job_category)
other_jobs_C <- job_category_counts %>%
  filter(n < 100) %>%
  pull(job_category)

data_science_grouped_C <- data_science %>%
  mutate(job_category_grouped_C = if_else(job_category %in% other_jobs_C, "Other", job_category))

plot7<- ggplot(data = data_science_grouped_C, aes(x = job_category_grouped_C, y = salary_in_usd, fill = job_category_grouped_C)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Data Science Salaries by Job Category",
       x = "Job Category",
       y = "Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  scale_y_continuous(labels = scales::label_comma())

plot_grid(plot6,plot7)
```

## The Effect of Company-Level Demands and Characteristics on Salary
```{r, echo=FALSE}
library(cowplot)

# summary statistics of average salary by work setting
work_salary <- data_science %>%
  group_by(work_setting) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))

# summary statistics of average salary by company size
company_salary <- data_science %>%
  group_by(company_size) %>%
  summarize(avg_salary = mean(salary_in_usd, na.rm = TRUE))


# Plot 3: Average Salary by Work Setting
plot1 <- ggplot(data = data_science, aes(x = work_setting, y = salary_in_usd, fill = work_setting)) +
  geom_violin() +
  theme_minimal() +
  labs(title = "Data Science Salaries by Work Setting",
       x = "Work Setting",
       y = "Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  legend.position = "none")+
  scale_y_continuous(labels = scales::label_comma())

# Plot 4: Average Salary by Company Size
plot2 <- ggplot(data = data_science, aes(x = factor(company_size, levels = c("S", "M", "L")), y = salary_in_usd, fill = factor(company_size, levels = c("S", "M", "L")))) +
  geom_violin() +
  theme_minimal() +
  labs(title = "Data Science Salaries by Company Size",
       x = "Company Size",
       y = "Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  legend.position = "none") +
  scale_x_discrete(labels = c("S" = "Small", "M" = "Medium", "L" = "Large"))+
  scale_y_continuous(labels = scales::label_comma())

# Arrange the plots in a grid
combined_plot <- plot_grid(plot1, plot2, ncol = 2)

# Print the combined plot
combined_plot
```

## The Effect of Personal Characteristics on Salary

```{r, echo=FALSE}
library(cowplot)

# Plot 1: Average Salary by Experience Level
plot3 <- ggplot(data = data_science, aes(x = experience_level, y = salary_in_usd, fill = experience_level)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Data Science Salaries by Experience Level",
       x = "Experience Level",
       y = "Avg Salary") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  legend.position = "none")


# Plot 2: Average Salary by Employment Type
plot4 <- ggplot(data = data_science, aes(x = employment_type, y = salary_in_usd, fill = employment_type)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Data Science Salaries by Employment Type",
       x = "Employment Type",
       y = "Salary (USD)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

# Arrange the plots in a grid
combined_plot2 <- plot_grid(plot3, plot4, ncol = 2)

# Print the combined plot
combined_plot2
```


## The Effect of Personal Characteristics on Salary
```{r, echo=FALSE}
library(tmap)
income <- read_csv(file = "data/median-income-by-country-2024.csv")

avg_salary_by_country <- data_science %>%
  group_by(company_location) %>%
  summarize(avg_salary_in_usd = mean(salary_in_usd))

avg_salary_by_country <- avg_salary_by_country %>%
  mutate(company_location = case_when(
    company_location == "Czech Republic" ~ "Czech Rep.",
    company_location == "South Korea" ~ "Korea",
    company_location == "Central African Republic" ~ "Central African Rep.",
    company_location == "Bosnia and Herzegovina" ~ "Bosnia and Herz.",
    TRUE ~ company_location
  ))


data("World")
world_map <- merge(avg_salary_by_country, income, by.x = "company_location", by.y = "country", all.x = TRUE) %>% mutate(normalized = avg_salary_in_usd/medianIncomeByCountry_meanIncome)

world_map <- merge(World, world_map, by.x = "name", by.y = "company_location", all.x = TRUE)

tm_shape(world_map) +
  tm_polygons(col = "avg_salary_in_usd", style = "jenks", palette = "Blues",
              main.title = "Average Salaries of Data Scientists by Country, 2020-2023",
              title = "Average Salaries by Country (USD), 2020-2023", legend.title = "Salary Range")
```

## The Effect of Personal Characteristics on Salary
```{r, echo = F}
tm_shape(world_map) +
  tm_polygons(col = "normalized", style = "jenks", palette = "Blues",
              main.title = "Average Salaries of Data Scientists by Country, 2020-2023",
              title = "Normalized Avg Salaries by Country (USD), 2020-2023", legend.title = "Salary Range")
```

